---
title: "choros evaluation"
author: "Amanda Mok"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, echo=F}
knitr::opts_chunk$set(fig.width=12, echo=F, warning=F)

library(ggplot2)
library(patchwork)

num_genes <- c(50, 75, 100, 125, 150, 175, 200)
main_terms <- c("transcript", "A", "P", "E", "d5", "d3", "genome_f5", "genome_f3")
```

#### Weinberg data

```{r weinberg_setup}
expt <- "weinberg"
expt_dir <- "."

load(file.path(expt_dir, paste0(expt, "_training_plots.Rda")))
load(file.path(expt_dir, paste0(expt, "_test_plots.Rda")))
load(file.path(expt_dir, paste0(expt, "_bam.Rda")))
load(file.path(expt_dir, paste0(expt, "_subsets.Rda")))

for(n in num_genes) {
  load(file.path(expt_dir, paste0(expt, "_coef_", n, ".Rda")))
}
```

```{r weinberg_subsets}
d5_d3$plot + ggtitle(paste(sub("_", "|", 
                               sub("d5_", "d5:", 
                                   sub("d3_", "d3:", subset_names))), 
                           collapse=" ; "))
```

Bias evaluation

```{r weinberg_bias, fig.height=6}
weinberg_training_plots
weinberg_test_plots
```

Consistency of regression coefficient terms

```{r weinberg_consistency_main, fig.height=30}
# main terms
weinberg_main_coef <- lapply(num_genes,
                          function(x) {
                            tmp_coef <- get(paste0("weinberg_coef_", x))
                            tmp_coef <- tmp_coef[!grepl("(Intercept)|:", names(tmp_coef))]
                            tmp_coef <- data.frame(term=names(tmp_coef),
                                                   value=tmp_coef,
                                                   num_genes=x,
                                                   type=NA,
                                                   stringsAsFactors=F, row.names=NULL)
                            for(y in main_terms) {
                              tmp_terms <- grepl(paste0("^", y), tmp_coef$term)
                              tmp_coef$term[tmp_terms] <- sub(y, "", tmp_coef$term[tmp_terms])
                              tmp_coef$type[tmp_terms] <- y
                            }
                            return(tmp_coef)
                          })
weinberg_main_coef <- do.call(rbind, weinberg_main_coef)
weinberg_main_plots <- lapply(main_terms,
                           function(x) {
                             tmp_coef <- subset(weinberg_main_coef, type==x)
                             ggplot(tmp_coef, aes(x=num_genes, y=value, col=term)) +
                               geom_line() + geom_point() + theme_bw() +
                               xlab("") + ylab(expr(beta)) + ggtitle(x) + theme(legend.position="none")
                           })
weinberg_main_plots[[length(main_terms)]] <- weinberg_main_plots[[length(main_terms)]] + 
  xlab("number of genes in training set")
wrap_plots(weinberg_main_plots, ncol=1)
```

```{r weinberg_consistency_interxn, fig.height=8}
# f5 interaction terms
weinberg_f5_interxn <- lapply(num_genes,
                           function(x) {
                             tmp_coef <- get(paste0("weinberg_coef_", x))
                             tmp_coef <- tmp_coef[grepl(":genome_f5", names(tmp_coef))]
                             tmp_coef <- data.frame(term=names(tmp_coef),
                                                    value=tmp_coef,
                                                    num_genes=x,
                                                    stringsAsFactors=F, row.names=NULL)
                             tmp_coef$d5 <- factor(sub("d5", "d5=", sub(":.*", "", tmp_coef$term)))
                             tmp_coef$term <- substr(tmp_coef$term, 
                                                     start=nchar(tmp_coef$term)-1,
                                                     stop=nchar(tmp_coef$term))
                             tmp_coef$end_nt <- substr(tmp_coef$term, start=1, stop=1) 
                             return(tmp_coef)
                           })
weinberg_f5_interxn <- do.call(rbind, weinberg_f5_interxn)
ggplot(weinberg_f5_interxn, aes(x=num_genes, y=value, col=term)) + geom_line() +
  geom_point(aes(shape=end_nt), size=2) + theme_bw() + facet_wrap(~d5, ncol=1) + 
  ggtitle("d5*f5 interaction terms") + xlab("number of genes in training set") + ylab(expr(beta))

# f3 interaction terms
weinberg_f3_interxn <- lapply(num_genes,
                           function(x) {
                             tmp_coef <- get(paste0("weinberg_coef_", x))
                             tmp_coef <- tmp_coef[grepl(":genome_f3", names(tmp_coef))]
                             tmp_coef <- data.frame(term=names(tmp_coef),
                                                    value=tmp_coef,
                                                    num_genes=x,
                                                    stringsAsFactors=F, row.names=NULL)
                             tmp_coef$d3 <- factor(sub("d3", "d3=", sub(":.*", "", tmp_coef$term)))
                             tmp_coef$term <- substr(tmp_coef$term, 
                                                     start=nchar(tmp_coef$term)-2,
                                                     stop=nchar(tmp_coef$term))
                             tmp_coef$end_nt <- substr(tmp_coef$term, 
                                                       start=nchar(tmp_coef$term), 
                                                       stop=nchar(tmp_coef$term)) 
                             return(tmp_coef)
                           })
weinberg_f3_interxn <- do.call(rbind, weinberg_f3_interxn)
ggplot(weinberg_f3_interxn, aes(x=num_genes, y=value, col=term)) + geom_line() +
  geom_point(aes(shape=end_nt), size=2) + theme_bw() + facet_wrap(~d3, ncol=1) + 
  ggtitle("d3*f3 interaction terms") + xlab("number of genes in training set") + ylab(expr(beta))
```

5' footprint recovery terms across digest lengths (trained on 150 genes)

```{r weinberg_d5f5_compare, fig.height=12}
weinberg_f5 <- unique(weinberg_main_coef$term[weinberg_main_coef$type=="genome_f5"])
weinberg_d5_ref <- unique(d5_d3_subsets$d5)[!(paste0("d5", unique(d5_d3_subsets$d5)) %in% names(weinberg_coef_150))]
weinberg_f5_ref <- sapply(weinberg_f5,
                       function(x) {
                         weinberg_main_coef$value[weinberg_main_coef$term==x &
                                                 weinberg_main_coef$type=="genome_f5" &
                                                 weinberg_main_coef$num_genes==150]
                       })

weinberg_d5f5_compare <- data.frame(term=weinberg_f5)
for(d5 in unique(d5_d3_subsets$d5)) {
  if(d5 == weinberg_d5_ref) {
    weinberg_d5f5_compare[, paste0("d5_", d5)] <- weinberg_f5_ref
  } else {
    tmp_compare <- sapply(weinberg_f5,
                          function(x) {
                            weinberg_f5_interxn$value[weinberg_f5_interxn$d5==paste0("d5=", d5) &
                                                     weinberg_f5_interxn$term==x &
                                                     weinberg_f5_interxn$num_genes==150]
                          })
    tmp_compare <- tmp_compare + weinberg_f5_ref
    weinberg_d5f5_compare[, paste0("d5_", d5)] <- tmp_compare
  }
}

weinberg_d5f5_plots <- expand.grid(sort(colnames(weinberg_d5f5_compare)[-1]), 
                                sort(colnames(weinberg_d5f5_compare)[-1]),
                                stringsAsFactors=F)
weinberg_d5f5_plots <- lapply(seq(nrow(weinberg_d5f5_plots)),
                           function(x) {
                             tmp_x <- weinberg_d5f5_plots$Var1[x]
                             tmp_y <- weinberg_d5f5_plots$Var2[x]
                             tmp_cor <- cor(weinberg_d5f5_compare[, tmp_x], weinberg_d5f5_compare[, tmp_y])
                             tmp_cor <- round(tmp_cor, 3)
                             tmp_plot <- ggplot(weinberg_d5f5_compare, 
                                                aes_string(x=tmp_x, y=tmp_y, col="term")) +
                               geom_point() + geom_abline(slope=1, intercept=0) + theme_bw() + 
                               theme(legend.position="none") + xlim(-1.5, 1) + ylim(-1.5, 1) +
                               xlab(sub("_", " = ", tmp_x)) + ylab(sub("_", " = ", tmp_y)) +
                               ggtitle(paste("cor =", tmp_cor))
                             return(tmp_plot)
                           })
wrap_plots(weinberg_d5f5_plots, byrow=T, ncol=length(unique(d5_d3_subsets$d5)))
```

3' footprint recovery terms across digest lengths (trained on 150 genes)

```{r weinberg_d3f3_compare}
weinberg_f3 <- unique(weinberg_main_coef$term[weinberg_main_coef$type=="genome_f3"])
weinberg_d3_ref <- unique(d5_d3_subsets$d3)[!(paste0("d3", unique(d5_d3_subsets$d3)) %in% names(weinberg_coef_150))]
weinberg_f3_ref <- sapply(weinberg_f3,
                       function(x) {
                         weinberg_main_coef$value[weinberg_main_coef$term==x &
                                                 weinberg_main_coef$type=="genome_f3" &
                                                 weinberg_main_coef$num_genes==150]
                       })

weinberg_d3f3_compare <- data.frame(term=weinberg_f3)
for(d3 in unique(d5_d3_subsets$d3)) {
  if(d3 == weinberg_d3_ref) {
    weinberg_d3f3_compare[, paste0("d3_", d3)] <- weinberg_f3_ref
  } else {
    tmp_compare <- sapply(weinberg_f3,
                          function(x) {
                            weinberg_f3_interxn$value[weinberg_f3_interxn$d3==paste0("d3=", d3) &
                                                     weinberg_f3_interxn$term==x &
                                                     weinberg_f3_interxn$num_genes==150]
                          })
    tmp_compare <- tmp_compare + weinberg_f3_ref
    weinberg_d3f3_compare[, paste0("d3_", d3)] <- tmp_compare
  }
}

weinberg_d3f3_plots <- expand.grid(sort(colnames(weinberg_d3f3_compare)[-1]), 
                                sort(colnames(weinberg_d3f3_compare)[-1]),
                                stringsAsFactors=F)
weinberg_d3f3_plots <- lapply(seq(nrow(weinberg_d3f3_plots)),
                           function(x) {
                             tmp_x <- weinberg_d3f3_plots$Var1[x]
                             tmp_y <- weinberg_d3f3_plots$Var2[x]
                             tmp_cor <- cor(weinberg_d3f3_compare[, tmp_x], weinberg_d3f3_compare[, tmp_y])
                             tmp_cor <- round(tmp_cor, 3)
                             tmp_plot <- ggplot(weinberg_d3f3_compare, 
                                                aes_string(x=tmp_x, y=tmp_y, col="term")) +
                               geom_point() + geom_abline(slope=1, intercept=0) + theme_bw() + 
                               theme(legend.position="none") + xlim(-1.3, 1) + ylim(-1.3, 1) +
                               xlab(sub("_", " = ", tmp_x)) + ylab(sub("_", " = ", tmp_y)) +
                               ggtitle(paste("cor =", tmp_cor))
                             return(tmp_plot)
                           })
wrap_plots(weinberg_d3f3_plots, byrow=T, ncol=length(unique(d5_d3_subsets$d3)))
```

Per-transcript footprint counts, before and after correction

```{r weinberg_transcript, fig.height=12, cache=T}
weinberg_transcripts <- levels(weinberg_bam$transcript)

weinberg_transcripts_counts <- data.frame(sapply(c("count", grep("correct", colnames(weinberg_bam), value=T)),
                                              function(x) {
                                                tmp_counts <- aggregate(formula(paste(x, "~ transcript")),
                                                                        data=weinberg_bam, FUN=sum)
                                                return(tmp_counts[match(weinberg_transcripts, 
                                                                        tmp_counts$transcript), x])
                                              }))
weinberg_transcripts_counts <- lapply(grep("correct", colnames(weinberg_bam), value=T),
                                   function(x) {
                                     data.frame(transcript=weinberg_transcripts,
                                                uncorrected=weinberg_transcripts_counts$count,
                                                corrected=weinberg_transcripts_counts[,x],
                                                num_genes=sub("correct_", "", x))
                                   })
weinberg_transcripts_counts <- do.call(rbind, weinberg_transcripts_counts)
levels(weinberg_transcripts_counts$num_genes) <- paste("trained on", 
                                                    levels(weinberg_transcripts_counts$num_genes), 
                                                    "genes")

ggplot(weinberg_transcripts_counts, aes(x=uncorrected, y=corrected)) + geom_point(alpha=0.5) + 
  geom_abline(slope=1, intercept=0, col="grey") + facet_wrap(~num_genes, ncol=1) + theme_bw() +
  ggtitle("All transcripts")

ggplot(weinberg_transcripts_counts, aes(x=uncorrected, y=corrected)) + geom_point(alpha=0.5) + 
  geom_abline(slope=1, intercept=0, col="grey") + facet_wrap(~num_genes, ncol=1) + theme_bw() +
  xlim(0, 50000) + ylim(0, 50000) + ggtitle("All transcripts, zoomed in")
```

Per-codon footprint counts, before and after correction

```{r weinberg_codon, fig.height=12, cache=T}
weinberg_codons <- aggregate(count ~ transcript + cod_idx, data=weinberg_bam, FUN=sum)

for(x in grep("correct", colnames(weinberg_bam), value=T)) {
  tmp_codons <- aggregate(formula(paste(x, "~ transcript + cod_idx")), data=weinberg_bam, FUN=sum)
  tmp_rows <- prodlim::row.match(weinberg_codons[, c("transcript", "cod_idx")],
                                 tmp_codons[, c("transcript", "cod_idx")])
  weinberg_codons[, x] <- tmp_codons[tmp_rows, x]
}

weinberg_codons_subsample <- weinberg_codons[sample.int(n=nrow(weinberg_codons), size=5e5), ]
weinberg_codons_subsample <- lapply(grep("correct", colnames(weinberg_bam), value=T),
                                 function(x) {
                                   data.frame(transcript=weinberg_codons_subsample$transcript,
                                              cod_idx=weinberg_codons_subsample$cod_idx,
                                              uncorrected=weinberg_codons_subsample$count,
                                              corrected=weinberg_codons_subsample[,x],
                                              num_genes=sub("correct_", "", x))
                                 })
weinberg_codons_subsample <- do.call(rbind, weinberg_codons_subsample)
levels(weinberg_codons_subsample$num_genes) <- paste("trained on", 
                                                  levels(weinberg_codons_subsample$num_genes),
                                                  "genes")

ggplot(weinberg_codons_subsample, aes(x=uncorrected, y=corrected)) + 
  geom_point(alpha=0.1) + geom_abline(slope=1, intercept=0, col="grey") + facet_wrap(~num_genes, ncol=1) + 
  theme_bw() + ggtitle("Random subset of 500,000 codon positions") + ylim(0, 50000)

ggplot(weinberg_codons_subsample, aes(x=uncorrected, y=corrected)) + 
  geom_point(alpha=0.1) + geom_abline(slope=1, intercept=0, col="grey") + facet_wrap(~num_genes, ncol=1) + 
  theme_bw() + ggtitle("Random subset of 500,000 codon positions, zoomed in") + 
  coord_cartesian(xlim=c(0,500), ylim=c(0, 2000))
```

Scaling factors

```{r weinberg_scaling}
weinberg_transcripts_counts$scaling <- with(weinberg_transcripts_counts, corrected/uncorrected)

weinberg_codon_scalings <- lapply(grep("correct", colnames(weinberg_bam), value=T),
                               function(x) {
                                 tmp_scaling <- data.frame(transcript=weinberg_codons$transcript,
                                                           cod_idx=weinberg_codons$cod_idx,
                                                           scaling=weinberg_codons[,x]/weinberg_codons$count,
                                                           num_genes=sub("correct_", "", x))
                                 tmp_scaling <- subset(tmp_scaling, weinberg_codons$count >=10)
                                 return(tmp_scaling)
                               })
weinberg_codon_scalings <- do.call(rbind, weinberg_codon_scalings)

(ggplot(weinberg_transcripts_counts, aes(x=num_genes, y=scaling, fill=num_genes)) + geom_violin() + 
    scale_y_log10() + theme_bw() + theme(legend.position="none") + geom_hline(yintercept=1) +
    xlab("") + ylab("scaling factor") + ggtitle("Per transcript") + theme(axis.text.x=element_blank())) / 
  (ggplot(weinberg_codon_scalings, aes(x=num_genes, y=scaling, fill=num_genes)) + geom_violin() + 
     scale_y_log10() + theme_bw() + theme(legend.position="none") + geom_hline(yintercept=1) +
     xlab("number of genes in training set") + ylab("scaling factor") + ggtitle("Per codon"))
```

Unestimatable regression coefficients, by number of genes in training set

```{r weinberg_no_coef}
weinberg_na_coef <- do.call(rbind,
                         lapply(num_genes,
                                function(x) {
                                  tmp_coef <- get(paste0("weinberg_coef_", x))
                                  tmp_na <- is.na(tmp_coef)
                                  if(sum(tmp_na)>0) {
                                    tmp_na <- data.frame(term=names(tmp_coef)[tmp_na],
                                                         num_genes=x,
                                                         stringsAsFactors=F)
                                    tmp_na$type <- sapply(tmp_na$term,
                                                          function(y) {
                                                            if(grepl(":", y)) {
                                                              tmp_term <- substr(strsplit(y, split=":")[[1]], 1, 2)
                                                            } else {
                                                              tmp_term <- sapply(main_terms,
                                                                                 function(z) {
                                                                                   grepl(z, y)
                                                                                 })
                                                              tmp_term <- main_terms[tmp_term]
                                                            }
                                                            return(paste(tmp_term, collapse="_"))
                                                          })
                                    return(tmp_na)
                                  } else {
                                    return(NULL)
                                  }
                                }))
weinberg_na_coef <- do.call(rbind, lapply(num_genes,
                                       function(x) {
                                         data.frame(type=unique(weinberg_na_coef$type),
                                                    count=sapply(unique(weinberg_na_coef$type),
                                                                 function(y) {
                                                                   sum(weinberg_na_coef$num_genes==x &
                                                                         weinberg_na_coef$type==y)
                                                                 }),
                                                    num_genes=x)
                                       }))
weinberg_na_coef$num_genes <- factor(weinberg_na_coef$num_genes)
ggplot(weinberg_na_coef, aes(x=num_genes, y=count, fill=type)) + geom_col() + theme_bw() + 
  xlab("number of genes in training set") + ylab("")
```

```{r weinberg_cleanup}
rm(list=grep("(weinberg)|(tmp)", ls(), value=T))
```
